Rationality, Evolution \& the Environment

### 2015-10-04 paper/paper.tex
minor edits

\begin{definition}[Rational choice from maximization of expected utility] Let $S$ be a (finite)
  set of states of the world and $A$ a (finite) set of actions. Given a probability
  distribution $P$ over $S$ and a utility function
  $u \mycolon S \times A \rightarrow \mathbb{R} $, an action $a^* \in A$ is a rational choice
  only if it maximizes the \emph{(subjective) expected utility}
  $\EU(a) = \sum_{s \in S} P(s) \ u(s, a)$: $a^* \in \text{argmax}_{a \in A} \EU(a)
  $.
\end{definition}

@@

It could be objected that, without additional constraints, the notion of rationality is likely
to collapse, as it seems possible to deem rational everything that is observed, given the
freedom to adjust beliefs and preferences freely. Of course, there are many ways in which
ascription of beliefs and preferences is constrained further by normative considerations of
rationality. \todo{speak about Dutch books here?} To begin with, if the DM's uncertainty is
about objective chance, it can be argued that subjective beliefs should conform (at least
roughly) to objective chance. When the DM's uncertainty cannot be anchored in an objective
notion of chance, then temporal consistency in his decision making should at least constrain
his subjective belief, else he fall prey to exploitation by Dutch books. Similarly, in certain
situations it may make sense to postulate on normative grounds that subjective preferences
should be oriented towards tracking objective utility, e.g., fitness. For instance, in their
classic work \myalert{(Alchian, Friedman)} argue that profit maximization is a reasonable
assumption for characterizing outcomes in a competitive market because only firms behaving in a
manner that is consistent with profit maximization will survive in the long run.\todo{@Paolo:
  please add references}

An alternative to descriptive vs.~normative views on rationality is \emph{adaptationism}
\citep[e.g.][]{HagenChater2012:Decision-Making}. \todo{more references} Adaptationism aims to
explain choice behavior by appeal to evolutionary considerations. The key idea is that we
should see decision makers (human and other life forms) as having evolved mechanisms for
choosing that were fitness enhancing (in the relevant domains of choice making, at the relevant
time of evolution). When it comes to preferences, this is clearly related to Alchian and
Friedman's argument that firms \emph{should} optimize that which makes them ``live,'' so to
speak. For biological agents, the argument would be that subjective preferences \emph{should}
track whatever determines objective fitness.

Taking inspiration from recent economic literature on the evolution of preference, we would
like to put this latter idea to the test. We present a formal argument intended to support two
points: (i) subjective preferences that differ from objective fitness can be sustained and even
be adaptive; this is brought to the foreground if we acknowledge that (ii) it is important to
take into account the potential mutual dependence of subjective preferences and subjective
beliefs in adaptationist approaches to decision making. More specifically, we show that
subjective, non-veridical preference representations based on a notion of regret can be
adaptive and even outperform objective, veridical representations of the actual fitness, when
agents, at least on occasion, lack the ability to form reasonable probabilistic beliefs, in
which case they hold imprecise probabilistic beliefs (CITE). The following Section~XYZ will
introduce the basic notions needed for this argument, namely regret, imprecise probabilistic
beliefs and security strategies.

The method by which we want to advance our argument is to integrate ideas from the recent
literature on preference evolution (CITE) with basic tenets of adaptationism: we are interested
in studying the evolutionary competition of choice mechanism, conceived here as pairs of
(general ways of) subjective preference and subjective belief
representation. \textcolor{red}{EXAND}. This method is further detailed in
Section~\ref{sec:model}.

\section{Regret, imprecise probabilistic beliefs, choice mechanisms}

\subsubsection{Regret theory} \label{sec:regreTheory}

>>>>

\begin{definition}[Rational choice from maximization of expected utility]
  \label{def:rationality}
  Let $S$ be a (finite) set of states of the world and $A$ a (finite) set of actions. Given a
  probability distribution $P$ over $S$ and a utility function
  $u \mycolon S \times A \rightarrow \mathbb{R} $, an action $a^* \in A$ is a rational choice
  only if it maximizes the \emph{(subjective) expected utility}
  $\EU(a) = \sum_{s \in S} P(s) \ u(s, a)$: $a^* \in \text{argmax}_{a \in A} \EU(a) $.
\end{definition}

@@

It could be objected that, without additional constraints, the notion of rationality is likely
to collapse, as it seems possible to deem rational everything that is observed, given the
freedom to adjust beliefs and preferences freely. Of course, there are many ways in which
ascription of beliefs and preferences is constrained further by normative considerations of
rationality. To begin with, if the DM's uncertainty is about objective chance, it can be argued
that subjective beliefs should conform (at least roughly) to objective chance. When the DM's
uncertainty cannot be anchored in an objective notion of chance, then temporal consistency in
his decision making should at least constrain his subjective beliefs, else he fall prey to
exploitation by Dutch books. Similarly, in certain situations it may make sense to postulate on
normative grounds that subjective preferences should be oriented towards tracking objective
utility, e.g., fitness. For instance, in their classic work \myalert{(Alchian, Friedman)} argue
that profit maximization is a reasonable assumption for characterizing outcomes in a
competitive market because only firms behaving in a manner that is consistent with profit
maximization will survive in the long run.\todo{@Paolo: please add references}

An alternative to descriptive vs.~normative views on rationality is \emph{adaptationism}
\citep[e.g.][]{HagenChater2012:Decision-Making}. \todo{more references} Adaptationism aims to
explain choice behavior by appeal to evolutionary considerations. The key idea is that decision
makers (human and other life forms) have evolved choice mechanisms which were fitness enhancing
(in the relevant domains of choice making, at the relevant time of evolution). When it comes to
preferences, this is clearly related to Alchian and Friedman's argument that firms
\emph{should} optimize that which makes them ``live,'' so to speak. For biological agents, the
argument would be that subjective preferences \emph{should} track whatever determines objective
fitness.

We would like to put this latter idea to the test. We present a formal argument intended to
support the conclusion that subjective preferences that differ from objective fitness can be
adaptive. This is brought to the foreground if we acknowledge the potential mutual dependence
of subjective preferences and subjective beliefs in adaptationist approaches to decision
making. More specifically, we show that subjective, non-veridical preference representations
based on a notion of regret
\citep[e.g.][]{Savage1951:The-theory-of-s,LoomesSugden1982:Regret-Theory:-}\todo{add
  references?}\ can be adaptive and even outperform objective, veridical representations of the
actual fitness, when agents, at least on occasion, hold imprecise probabilistic beliefs
(CITE).

The method by which we want to advance our argument is to integrate ideas from the recent
literature on preference evolution (CITE) with basic tenets of adaptationism: we are interested
in studying the evolutionary competition of choice mechanism, conceived here as pairs of
subjective preference and subjective belief representation. Evolutionary competition between
choice mechanisms is driven by the expected objective fitness that each choice mechanisms
accrues when put to use repeatedly in a range of choice situations. We focus on the more
complex case of multi-agent decision making first, i.e., game situations. Solitary decision
making is then easy to deal with subsequently.

The following Section~XYZ will introduce the basic notions needed in the remainder. Section~XYZ
details our method of using ``meta-games''. \dots

\section{Basic notions}
\label{sec:basic-notions}

A strategic game is a tuple $ G=\tuple{ N, (A_i , \pi_i)_{i \in N} }$ with $N$ a set of
players, $A_i$ a set of acts and $\pi_i$ a payoff function for each player $i \in N$. The
classical notion of rationality from Definition~\ref{def:rationality} also applies to
games. Let $A_{-i} = \bigtimes A_j$ for all $j \neq i$ be the set of all profiles of acts for
all players excepts $i$. Then player $i$ has a belief $\Delta(A_{-i})$ about the behavior of
his opponents.

\subsection{Regret theory} \label{sec:regreTheory}
